---
title: "Home Credit - Data Preparation and Feature Engineering"
format:
  html:
    toc: true
    code-fold: show
execute:
  warning: false
  message: false
---

# Reusable Train/Test Feature Engineering Script

```{r}
library(data.table)
library(dplyr)
library(tidyr)
```

## 1) Utility helpers

```{r}
# Safe divide to avoid Inf/NaN in ratio features
safe_divide <- function(num, den, default = NA_real_) {
  out <- ifelse(!is.na(den) & den != 0, num / den, default)
  return(out)
}

# Replace infinite values after ratio operations
replace_infinite <- function(df, replacement = NA_real_) {
  num_cols <- names(df)[sapply(df, is.numeric)]
  for (col in num_cols) {
    x <- df[[col]]
    x[is.infinite(x)] <- replacement
    df[[col]] <- x
  }
  df
}
```

## 2) Application cleaning (train/test reusable)

```{r}
clean_application_core <- function(df) {
  # 1) DAYS_EMPLOYED anomaly from EDA: 365243 placeholder
  df <- df %>%
    mutate(
      DAYS_EMPLOYED_ANOM = as.integer(DAYS_EMPLOYED == 365243),
      DAYS_EMPLOYED = ifelse(DAYS_EMPLOYED == 365243, NA_real_, DAYS_EMPLOYED)
    )

  # 2) Missing indicators for EXT_SOURCE variables
  for (v in c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")) {
    if (v %in% names(df)) {
      df[[paste0(v, "_MISS")]] <- as.integer(is.na(df[[v]]))
    }
  }

  # 3) Convert negative day fields to positive interpretable units
  if ("DAYS_BIRTH" %in% names(df)) {
    df$AGE_YEARS <- abs(df$DAYS_BIRTH) / 365.25
  }
  if ("DAYS_EMPLOYED" %in% names(df)) {
    df$EMPLOYED_YEARS <- abs(df$DAYS_EMPLOYED) / 365.25
  }

  # 4) Other known issue from EDA: DAYS_LAST_PHONE_CHANGE weird negatives
  if ("DAYS_LAST_PHONE_CHANGE" %in% names(df)) {
    df <- df %>% mutate(DAYS_LAST_PHONE_CHANGE = ifelse(DAYS_LAST_PHONE_CHANGE > 0, NA, DAYS_LAST_PHONE_CHANGE))
  }

  df
}
```

## 3) Fit/train-time statistics (store and reuse on test)

```{r}
fit_preprocess_stats <- function(train_df) {
  stats <- list()

  # Medians from TRAIN only
  ext_vars <- intersect(c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3"), names(train_df))
  stats$ext_medians <- sapply(ext_vars, function(v) median(train_df[[v]], na.rm = TRUE))

  # Binning thresholds from TRAIN only
  if ("AGE_YEARS" %in% names(train_df)) {
    stats$age_breaks <- unique(quantile(train_df$AGE_YEARS, probs = c(0, .2, .4, .6, .8, 1), na.rm = TRUE))
  }
  if ("AMT_INCOME_TOTAL" %in% names(train_df)) {
    stats$income_breaks <- unique(quantile(train_df$AMT_INCOME_TOTAL, probs = c(0, .2, .4, .6, .8, 1), na.rm = TRUE))
  }

  stats
}
```

## 4) Apply imputations + engineered features

```{r}
apply_preprocess_and_engineer <- function(df, stats) {
  # Impute EXT_SOURCE with TRAIN medians
  if (!is.null(stats$ext_medians)) {
    for (v in names(stats$ext_medians)) {
      if (v %in% names(df)) {
        df[[v]][is.na(df[[v]])] <- stats$ext_medians[[v]]
      }
    }
  }

  # Demographic + stability features
  df <- df %>%
    mutate(
      # Employment-to-age proportion
      EMPLOYED_AGE_RATIO = safe_divide(EMPLOYED_YEARS, AGE_YEARS),

      # External source summary score
      EXT_SOURCE_MEAN = rowMeans(select(., any_of(c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3"))), na.rm = TRUE)
    )

  # Financial ratios (common lending ratios)
  df <- df %>%
    mutate(
      CREDIT_INCOME_RATIO = safe_divide(AMT_CREDIT, AMT_INCOME_TOTAL),
      ANNUITY_INCOME_RATIO = safe_divide(AMT_ANNUITY, AMT_INCOME_TOTAL),
      CREDIT_ANNUITY_RATIO = safe_divide(AMT_CREDIT, AMT_ANNUITY),
      INCOME_PER_PERSON = safe_divide(AMT_INCOME_TOTAL, CNT_FAM_MEMBERS),
      GOODS_CREDIT_DIFF = AMT_CREDIT - AMT_GOODS_PRICE,
      GOODS_CREDIT_RATIO = safe_divide(AMT_GOODS_PRICE, AMT_CREDIT),
      # Approximate loan-to-value proxy when goods price available
      LOAN_TO_VALUE_PROXY = safe_divide(AMT_CREDIT, AMT_GOODS_PRICE)
    )

  # Interaction terms
  df <- df %>%
    mutate(
      AGE_X_CREDIT_INCOME = AGE_YEARS * CREDIT_INCOME_RATIO,
      EMPLOYED_X_ANNUITY_INCOME = EMPLOYED_YEARS * ANNUITY_INCOME_RATIO
    )

  # Binned variables using TRAIN thresholds
  if (!is.null(stats$age_breaks) && length(stats$age_breaks) > 2) {
    df$AGE_BIN <- cut(df$AGE_YEARS, breaks = stats$age_breaks, include.lowest = TRUE, ordered_result = TRUE)
  }
  if (!is.null(stats$income_breaks) && length(stats$income_breaks) > 2) {
    df$INCOME_BIN <- cut(df$AMT_INCOME_TOTAL, breaks = stats$income_breaks, include.lowest = TRUE, ordered_result = TRUE)
  }

  df <- replace_infinite(df)
  df
}
```

## 5) Supplementary aggregation functions (applicant level: SK_ID_CURR)

```{r}
aggregate_bureau <- function(bureau_df) {
  bureau_df %>%
    group_by(SK_ID_CURR) %>%
    summarise(
      BUREAU_CREDIT_COUNT = n(),
      BUREAU_ACTIVE_COUNT = sum(CREDIT_ACTIVE == "Active", na.rm = TRUE),
      BUREAU_CLOSED_COUNT = sum(CREDIT_ACTIVE == "Closed", na.rm = TRUE),
      BUREAU_OVERDUE_SUM = sum(CREDIT_DAY_OVERDUE, na.rm = TRUE),
      BUREAU_DEBT_SUM = sum(AMT_CREDIT_SUM_DEBT, na.rm = TRUE),
      BUREAU_CREDIT_SUM = sum(AMT_CREDIT_SUM, na.rm = TRUE),
      BUREAU_DEBT_RATIO = safe_divide(BUREAU_DEBT_SUM, BUREAU_CREDIT_SUM, default = 0),
      .groups = "drop"
    )
}

aggregate_previous_application <- function(prev_df) {
  prev_df %>%
    group_by(SK_ID_CURR) %>%
    summarise(
      PREV_APP_COUNT = n(),
      PREV_APPROVED_COUNT = sum(NAME_CONTRACT_STATUS == "Approved", na.rm = TRUE),
      PREV_REFUSED_COUNT = sum(NAME_CONTRACT_STATUS == "Refused", na.rm = TRUE),
      PREV_APPROVAL_RATE = safe_divide(PREV_APPROVED_COUNT, PREV_APP_COUNT, default = 0),
      PREV_REFUSAL_RATE = safe_divide(PREV_REFUSED_COUNT, PREV_APP_COUNT, default = 0),
      .groups = "drop"
    )
}

aggregate_installments <- function(inst_df) {
  inst_df %>%
    mutate(
      DAYS_LATE = DAYS_ENTRY_PAYMENT - DAYS_INSTALMENT,
      LATE_FLAG = as.integer(DAYS_LATE > 0)
    ) %>%
    group_by(SK_ID_CURR) %>%
    summarise(
      INSTAL_COUNT = n(),
      INSTAL_LATE_COUNT = sum(LATE_FLAG, na.rm = TRUE),
      INSTAL_LATE_PCT = safe_divide(INSTAL_LATE_COUNT, INSTAL_COUNT, default = 0),
      INSTAL_DAYS_LATE_MEAN = mean(DAYS_LATE[DAYS_LATE > 0], na.rm = TRUE),
      INSTAL_PAYMENT_MEAN = mean(AMT_PAYMENT, na.rm = TRUE),
      INSTAL_PAYMENT_TREND = safe_divide(
        mean(AMT_PAYMENT[NUM_INSTALMENT_NUMBER > median(NUM_INSTALMENT_NUMBER, na.rm = TRUE)], na.rm = TRUE),
        mean(AMT_PAYMENT[NUM_INSTALMENT_NUMBER <= median(NUM_INSTALMENT_NUMBER, na.rm = TRUE)], na.rm = TRUE),
        default = 1
      ),
      .groups = "drop"
    )
}
```

## 6) Join all aggregated supplementary features

```{r}
join_supplementary_features <- function(app_df, bureau_agg, prev_agg, inst_agg) {
  app_df %>%
    left_join(bureau_agg, by = "SK_ID_CURR") %>%
    left_join(prev_agg, by = "SK_ID_CURR") %>%
    left_join(inst_agg, by = "SK_ID_CURR")
}
```

## 7) Train/Test consistency function

```{r}
align_train_test_columns <- function(train_df, test_df, target_col = "TARGET") {
  train_cols <- setdiff(names(train_df), target_col)
  test_cols <- names(test_df)

  # Add missing cols to test
  missing_in_test <- setdiff(train_cols, test_cols)
  for (col in missing_in_test) {
    test_df[[col]] <- NA
  }

  # Remove extra cols from test
  extra_in_test <- setdiff(test_cols, train_cols)
  if (length(extra_in_test) > 0) {
    test_df <- test_df %>% select(-all_of(extra_in_test))
  }

  # Reorder test to match train feature order
  test_df <- test_df %>% select(all_of(train_cols))

  list(train = train_df, test = test_df)
}
```

## 8) End-to-end pipeline callable for train and test

```{r}
process_train_test <- function(app_train, app_test, bureau, prev_app, installments) {
  # Clean base application
  train_clean <- clean_application_core(app_train)
  test_clean <- clean_application_core(app_test)

  # Fit train-only stats and apply to both
  stats <- fit_preprocess_stats(train_clean)
  train_feat <- apply_preprocess_and_engineer(train_clean, stats)
  test_feat <- apply_preprocess_and_engineer(test_clean, stats)

  # Aggregate supplementary data once
  bureau_agg <- aggregate_bureau(bureau
