---
title: "Home Credit Default Risk - Comprehensive Exploratory Data Analysis"
author: "Phan Chung"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

# Introduction and Setup

This notebook addresses several foundational exploratory data analysis questions for the Home Credit Default Risk competition before moving to modeling. 

We will explore:
1. Target variable distribution and baseline accuracy
2. Key predictors and their relationships with the target
3. Missing data patterns and imputation strategies
4. Data quality issues, anomalies, and outliers
5. Data transformations needed for modeling
6. Aggregating and joining supplementary transactional data

## Library Loading

```{r setup}
library(tidyverse)
library(data.table)
library(ggplot2)
library(corrplot)
library(skimr)
library(scales)
library(gridExtra)

# Set visualization theme
theme_set(theme_minimal(base_size = 12))
```

## Data Loading

Let's load the primary application files.

```{r load-data}
# Using data.table::fread for fast loading
train <- fread("application_train.csv")
test <- fread("application_test.csv")

cat("Training set dimensions:", dim(train)[1], "rows,", dim(train)[2], "columns\n")
cat("Test set dimensions:", dim(test)[1], "rows,", dim(test)[2], "columns\n")
```

---

# 1. Target Variable Analysis

## Distribution and Imbalance

*Question: Explore the target variable in application_{train|test}.csv. Is the data unbalanced with respect to the target? What would the accuracy be for a simple model consisting in a majority class classifier?*

```{r target-exploration}
# Check target distribution
target_counts <- table(train$TARGET)
target_props <- prop.table(target_counts)

# Create a summary dataframe
target_summary <- data.frame(
  Class = c("0 (Repaid)", "1 (Default)"),
  Count = as.numeric(target_counts),
  Percentage = as.numeric(target_props) * 100
)

print(target_summary)

# Visualize the imbalance
ggplot(target_summary, aes(x = Class, y = Count, fill = Class)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("steelblue", "darkred")) +
  labs(title = "Target Variable Distribution",
       subtitle = "Severe Class Imbalance Detected",
       x = "Target Class",
       y = "Number of Applicants") +
  theme(legend.position = "none")
```

## Baseline Majority Class Classifier

**Interpretation:**
Yes, the data is **highly imbalanced**. Approximately 91.9% of loans are repaid (Class 0) and only 8.1% default (Class 1). This is an imbalance ratio of roughly 11.4:1.

If we built a "majority class classifier" - a naive model that simply predicts "0 (Repaid)" for every single applicant - this model would achieve an **accuracy of 91.92%**. 

**Why this is a problem:**
Because of this extreme imbalance, pure accuracy is a terrible metric for this project. A useless model (predicting 0 for everyone) achieves 92% accuracy! We must use **Area Under the ROC Curve (AUC)** as our evaluation metric instead, as it measures the model's ability to discriminate between the two classes regardless of their distribution. We will also need to consider techniques like class weighting or SMOTE during modeling.

---

# 2. Key Predictors Exploration

*Question: Explore the relationship between target and predictors, looking for potentially strong predictors that could be included later in a model. Which are most predictive of the target? Visualize and interpret these relationships.*

Based on domain knowledge and previous competition solutions, the external source scores from credit bureaus are typically the strongest predictors. Let's examine these and a few structural variables.

## External Credit Scores

```{r ext-sources}
# Select the external sources and target
ext_data <- train %>% 
  select(TARGET, EXT_SOURCE_1, EXT_SOURCE_2, EXT_SOURCE_3) %>%
  mutate(TARGET = as.factor(TARGET))

# Correlation matrix for external sources
cor_data <- train %>% 
  select(TARGET, EXT_SOURCE_1, EXT_SOURCE_2, EXT_SOURCE_3) %>%
  drop_na()
  
cat("Correlations with TARGET:\n")
cor(cor_data)[1,]

# Create density plots for external sources by Target
p1 <- ggplot(ext_data, aes(x = EXT_SOURCE_1, fill = TARGET)) + 
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("steelblue", "darkred"), labels = c("Repaid", "Default")) +
  labs(title = "EXT_SOURCE_1 Distribution") +
  theme(legend.position = "none")

p2 <- ggplot(ext_data, aes(x = EXT_SOURCE_2, fill = TARGET)) + 
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("steelblue", "darkred"), labels = c("Repaid", "Default")) +
  labs(title = "EXT_SOURCE_2 Distribution") +
  theme(legend.position = "none")

p3 <- ggplot(ext_data, aes(x = EXT_SOURCE_3, fill = TARGET)) + 
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("steelblue", "darkred"), labels = c("Repaid", "Default")) +
  labs(title = "EXT_SOURCE_3 Distribution")

grid.arrange(p1, p2, p3, ncol = 3, widths = c(1, 1, 1.4))
```

**Interpretation:**
The `EXT_SOURCE_*` variables represent normalized scores from external data sources (likely credit bureaus). All three show a clear negative correlation with default risk: **lower external scores indicate a much higher probability of default**. The distributions for defaulters (red) are shifted significantly to the left compared to those who repaid (blue). These are exceptionally strong predictors that must be included in any model.

## Demographic and Age Features

Let's look at applicant age (`DAYS_BIRTH`). Note: Home Credit stores dates as negative days relative to the current application. We'll convert to positive years.

```{r age-relationship}
# Convert DAYS_BIRTH to Age in Years
age_data <- train %>%
  select(TARGET, DAYS_BIRTH) %>%
  mutate(
    Age_Years = as.integer(abs(DAYS_BIRTH) / 365),
    Age_Group = cut(Age_Years, breaks = seq(20, 70, by=5))
  )

# Calculate default rate by age group
age_defaults <- age_data %>%
  group_by(Age_Group) %>%
  summarise(
    Total = n(),
    Defaults = sum(TARGET),
    Default_Rate = (Defaults / Total) * 100
  ) %>%
  drop_na()

# Plot
ggplot(age_defaults, aes(x = Age_Group, y = Default_Rate)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(Default_Rate, 1), "%")), vjust = -0.5) +
  labs(title = "Default Rate by Applicant Age",
       subtitle = "Risk clearly decreases as applicants get older",
       x = "Age Group (Years)",
       y = "Default Rate (%)")
```

**Interpretation:**
There is a striking, near-linear relationship between age and default risk. The youngest applicants (20-25 years) default at a rate of 12.3%, while the oldest (65-70 years) default at only 3.7%. Age (`DAYS_BIRTH`) is highly predictable of target.

---

# 3. Missing Data Analysis

*Question: Explore the scope of missing data in application_{train|test}.csv and come up with possible solutions. Remove rows? Remove columns? Impute? Explain your proposed solution.*

```{r missing-data}
# Calculate missing data percentages for all columns
missing_counts <- colSums(is.na(train))
missing_df <- data.frame(
  Feature = names(missing_counts),
  Missing_Count = missing_counts,
  Missing_Pct = (missing_counts / nrow(train)) * 100
) %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Pct))

cat("Number of columns with missing data:", nrow(missing_df), "out of", ncol(train), "\n\n")
cat("Top 15 columns with most missing data:\n")
print(head(missing_df, 15))

# Plot missing value distribution
ggplot(missing_df, aes(x = Missing_Pct)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Missing Data Across Features",
       x = "Percentage of Data Missing",
       y = "Number of Features")
```

## Proposed Missing Data Strategy

**Analysis:**
The missing data problem is massive. 67 columns (out of 122) have some missing data. Almost 50 features have >40% missing data. Notably, `EXT_SOURCE_
