---
title: "Home Credit Modeling Notebook"
author: "Phan Chung"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

# Modeling Overview

This notebook builds a practical modeling workflow for Home Credit Default Risk with a strong focus on **out-of-sample AUC**, because the target is highly imbalanced (~8% default rate). 

The notebook covers:

1. Baseline benchmark (majority class and simple logistic regression)
2. Candidate model comparison with 3-fold cross-validation
3. Class imbalance strategies (none vs. SMOTE vs. downsampling)
4. Randomized hyperparameter search for the best model
5. Impact of supplementary features
6. Kaggle submission generation

To keep runtime manageable, hyperparameter tuning uses a 5K sample, 3-fold CV, and randomized search with ~20 iterations.

---

# 1) Setup

```{r setup}
library(data.table)
library(dplyr)
library(tidyr)
library(caret)
library(pROC)
library(xgboost)
library(randomForest)
library(ROSE)        # For ovun.sample (sampling strategies)
library(ggplot2)

set.seed(42)

cat("✅ Packages loaded\n")
```

---

# 2) Load Data

We use processed train/test data if available, because it already includes engineered and supplementary features.

```{r load-data}
# Prefer processed data from your preparation notebook
train_path_processed <- "application_train_processed.csv"
test_path_processed <- "application_test_processed.csv"

if (file.exists(train_path_processed) && file.exists(test_path_processed)) {
  train <- fread(train_path_processed)
  test <- fread(test_path_processed)
  cat("✅ Loaded processed train/test data\n")
} else if (file.exists("application_train.csv") && file.exists("application_test.csv")) {
  train <- fread("application_train.csv")
  test <- fread("application_test.csv")
  cat("⚠️ Processed files not found; loaded raw application files\n")
} else if (file.exists("Data/application_train.csv") && file.exists("application_test.csv")) {
  train <- fread("Data/application_train.csv")
  test <- fread("application_test.csv")
  cat("⚠️ Processed files not found; loaded Data/application_train.csv + application_test.csv\n")
} else if (file.exists("../application_train.csv") && file.exists("../application_test.csv")) {
  train <- fread("../application_train.csv")
  test <- fread("../application_test.csv")
  cat("⚠️ Loaded train/test from parent folder\n")
} else if (file.exists("../Data/application_train.csv") && file.exists("../application_test.csv")) {
  train <- fread("../Data/application_train.csv")
  test <- fread("../application_test.csv")
  cat("⚠️ Loaded train/test from parent Data folder\n")
} else {
  stop("Could not find train/test files. Check file paths in the load-data chunk.")
}

cat("Train shape:", nrow(train), "x", ncol(train), "\n")
cat("Test shape:", nrow(test), "x", ncol(test), "\n")
```

---

# 3) Establish Baseline Performance

A baseline helps us understand whether more complex models add real value.

## 3.1 Majority Class Baseline

```{r baseline-majority}
# Check target balance
class_counts <- table(train$TARGET)
class_props <- prop.table(class_counts)

print(class_counts)
print(round(class_props, 4))

# Majority class prediction: always predict 0
majority_class <- as.numeric(names(which.max(class_counts)))
y_true <- train$TARGET
y_pred_majority <- rep(majority_class, length(y_true))
y_prob_majority <- rep(0, length(y_true))  # constant score

baseline_accuracy <- mean(y_pred_majority == y_true)
baseline_auc <- as.numeric(auc(y_true, y_prob_majority))

cat("Majority baseline accuracy:", round(baseline_accuracy, 4), "\n")
cat("Majority baseline AUC:", round(baseline_auc, 4), "\n")
```

**Interpretation:** High accuracy here is expected due to class imbalance and is not useful. AUC of ~0.50 confirms no ranking ability.

## 3.2 Simple Logistic Regression Baseline

To keep runtime fast, we use a small predictor subset on a 5K sample.

```{r baseline-logistic}
# Sample for quick baseline
sample_idx <- sample(1:nrow(train), min(5000, nrow(train)))
train_small <- train[sample_idx, ]

# Select a minimal predictor set if available
candidate_vars <- c(
  "EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3",
  "AMT_CREDIT", "AMT_INCOME_TOTAL", "AMT_ANNUITY",
  "DAYS_BIRTH", "DAYS_EMPLOYED"
)

use_vars <- intersect(candidate_vars, names(train_small))

logit_df <- train_small %>%
  select(all_of(c("TARGET", use_vars)))

# Basic NA handling for logistic model
for (col in names(logit_df)) {
  if (is.numeric(logit_df[[col]])) {
    logit_df[[col]][is.na(logit_df[[col]])] <- median(logit_df[[col]], na.rm = TRUE)
  }
}

# Fit logistic model
formula_logit <- as.formula(paste("TARGET ~", paste(use_vars, collapse = " + ")))
logit_model <- glm(formula_logit, data = logit_df, family = binomial())

logit_probs <- predict(logit_model, newdata = logit_df, type = "response")
logit_preds <- ifelse(logit_probs > 0.5, 1, 0)

logit_accuracy <- mean(logit_preds == logit_df$TARGET)
logit_auc <- as.numeric(auc(logit_df$TARGET, logit_probs))

cat("Logistic baseline accuracy:", round(logit_accuracy, 4), "\n")
cat("Logistic baseline AUC:", round(logit_auc, 4), "\n")
```

---

# 4) Candidate Model Comparison (3-Fold CV)

We compare model families by **cross-validated out-of-sample AUC**.

```{r prep-model-data}
# Remove identifier and split target
drop_cols <- intersect(c("SK_ID_CURR"), names(train))
model_df <- train %>% select(-all_of(drop_cols))

# Separate target
y <- model_df$TARGET
X <- model_df %>% select(-TARGET)

# Basic preprocessing for mixed models:
# 1) Encode character/factor as integers
for (col in names(X)) {
  if (!is.numeric(X[[col]])) {
    X[[col]] <- as.numeric(as.factor(X[[col]]))
  }
}

# 2) Replace NA with sentinel for tree models
X[is.na(X)] <- -999

# For speed in exploration: use sample
cv_idx <- sample(1:nrow(X), min(10000, nrow(X)))
X_cv <- X[cv_idx, ]
y_cv <- y[cv_idx]

# 3-fold stratified folds
folds <- createFolds(y_cv, k = 3, list = TRUE, returnTrain = FALSE)

cat("CV sample rows:", nrow(X_cv), "\n")
```

```{r cv-helper-functions}
cv_auc_logistic <- function(X_df, y_vec, folds) {
  aucs <- c()
  for (i in seq_along(folds)) {
    test_idx <- folds[[i]]
    train_idx <- setdiff(seq_len(nrow(X_df)), test_idx)

    X_train <- X_df[train_idx, ]
    y_train <- y_vec[train_idx]
    X_test <- X_df[test_idx, ]
    y_test <- y_vec[test_idx]

    # Median imputation for logistic
    for (col in names(X_train)) {
      med <- median(X_train[[col]], na.rm = TRUE)
      X_train[[col]][is.na(X_train[[col]])] <- med
      X_test[[col]][is.na(X_test[[col]])] <- med
    }

    train_df <- cbind(TARGET = y_train, X_train)
    model <- glm(TARGET ~ ., data = train_df, family = binomial())
    probs <- predict(model, newdata = X_test, type = "response")

    aucs <- c(aucs, as.numeric(auc(y_test, probs)))
  }
  aucs
}

cv_auc_rf <- function(X_df, y_vec, folds, ntree = 200) {
  aucs <- c()
  for (i in seq_along(folds)) {
    test_idx <- folds[[i]]
    train_idx <- setdiff(seq_len(nrow(X_df)), test_idx)

    X_train <- X_df[train_idx, ]
    y_train <- as.factor(y_vec[train_idx])
    X_test <- X_df[test_idx, ]
    y_test <- y_vec[test_idx]

    rf_model <- randomForest(
      x = X_train,
      y = y_train,
      ntree = ntree,
      importance = FALSE
    )

    probs <- predict(rf_model, X_test, type = "prob")[, "1"]
    aucs <- c(aucs, as.numeric(auc(y_test, probs)))
  }
  aucs
}

cv_auc_xgb <- function(X_df, y_vec, folds, params = list()) {
  aucs <- c()

  # Default fast params
  default_params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = 6,
    eta = 0.1,
    subsample = 0.8,
    colsample_bytree = 0.8
  )
  params <- modifyList(default_params, params)

  for (i in seq_along(folds)) {
    test_idx <- folds[[i]]
    train_idx <- setdiff(seq_len(nrow(X_df)), test_idx)

    X_train <- as.matrix(X_df[train_idx, ])
    y_train <- y_vec[train_idx]
    X_test <- as.matrix(X_df[test_idx, ])
    y_test <- y_vec[test_idx]

    dtrain <- xgb.DMatrix(data = X_train, label = y_train)
    dtest <- xgb.DMatrix(data = X_test)

    xgb_model <- xgb.train(
      params = params,
      data = dtrain,
      nrounds = 120,
      verbose = 0
    )

    probs <- predict(xgb_model, dtest)
    aucs <- c(aucs, as.numeric(auc(y_test, probs)))
  }
  aucs
}
```

```{r candidate-models-cv}
# Logistic Regression
auc_logit <- cv_auc_logistic(X_cv, y_cv, folds)

# Random Forest
auc_rf <- cv_auc_rf(X_cv, y_cv, folds, ntree = 200)

# XGBoost (no imbalance adjustment yet)
auc_xgb <- cv_auc_xgb(X_cv, y_cv, folds)

model_compare <- tibble(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  Mean_AUC = c(mean(auc_logit), mean(auc_rf), mean(auc_xgb)),
  SD_AUC = c(sd(auc_logit), sd(auc_rf), sd(auc_xgb))
)

model_compare %>% arrange(desc(Mean_AUC))
```

```{r model-comparison-plot}
ggplot(model_compare, aes(x = reorder(Model, Mean_AUC), y = Mean_AUC)) +
  geom_col(fill = "steelblue", alpha = 0.8
